Supplemental Materials: \emph{Simultaneous segmentation and registration of diffusion MR images of the brain driven by active-contours}

### 2015-03-18 Publications/2015-NeuroImage/04-methods.tex
final fixes
update supplementals
improving to second round

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \int_{\Omega} \underset{l}{\sum} \mdist{f'}{l} \,d\vec{r} \notag\\
  &+ \int_{\Omega} [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}
  \Big\}.
  \label{eq:gradient_descent}
  \end{align}

We apply a discretized interpretation of \eqref{eq:shape_gradients} to compute
  the data term in \eqref{eq:gradient_descent} using shape-gradients
  \citep{herbulot_segmentation_2006}, and ultimately avoid level sets.
We refer the reader to the appendix \autoref{app:shape_priors} in order to fill in the gap
  between \eqref{eq:gradient_descent} and the following expression:

@@

  \begin{equation}
  sWI = \frac{1}{P} \sum\limits_p^P \frac{1}{A_p} \sum\limits_i^{N_p} a_i\,\|
  \vec{v}_i - \hat{\vec{v}}_i \|
  \label{eq:swindex}
  \end{equation}
%
  where $\vec{v}_i$ are the locations of the $N_p$ vertices in each $p \in \{1, \dots P\}$
  prior, $A_p$ the total area of surface $p$, and $\hat{\vec{v}}_i$ is the location
  recovered corresponding to the vertex $\vec{v}_i$.

@@

\paragraph*{Simulated phantoms}%
\label{sec:digital_phantoms}
We proved the concept on simplistic digital phantoms which we will name after their
  appearance as ``box'', ``ball'', ``L-shape'', and ``gyrus'' (\autoref{fig:phantom},
  boxes A and B).
Using \emph{phantomas} \citep{caruyer_phantomas_2014}, we simulated
  \gls*{t1} (TE/TR=10/1500ms) and \gls*{t2} images (TE/TR=90/5000ms)
  corresponding to each phantom type, with two resolutions each
  ($1.0mm$ and $2.0mm$ isotropic).
Simulations were corrupted with rician noise for a \gls*{snr} of 300.0.
The reference surfaces were extracted at the higher resolution (1.0mm isotropic),
  using \texttt{mri\_tessellate}\footnote{A marching-cubes algorithm shipped with 
  \emph{FreeSurfer} \citep{fischl_freesurfer_2012}}.

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/figure03}
	\caption{A. The ``cortex'' phantom is a spherical shape with two sulci and an
	  outer crust resembling the cortical folding (left).
	The model is used to generate \gls*{t1} and \gls*{t2} images after warping the
	  contours using a random and plausible transformation $U_{true}^{-1}$ (right).
	B. Visual assessment of the results on the low resolution sets:
	  ``gyrus'' (top-left), ``L-shape'' (top-right), ``ball'' (bottom-left),
	  and ``box'' at (bottom-right).
	In yellow color, the recovered contours after registration are represented.
	Our method showed high accuracy, as demonstrates almost exact location of the registered
    contours with respect to their ground truth position depicted in green.
	Partial volume effect turns segmentation of the sulci a challenging problem with voxel-wise
	  clustering methods, but it is successfully segmented with our method.
	C. Quantitative evaluation of registration error in terms of average Hausdorff distance of
	  surfaces at high (left) and low (right) resolutions, demonstrating that the error is
	  consistently below the voxel size.
	  }\label{fig:phantom}
\end{figure*}

>>>>

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \notag\\
  &+ \int_{\Omega} [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}
  \Big\}.
  \label{eq:gradient_descent}
  \end{align}

We apply a discretized interpretation of \eqref{eq:shape_gradients} to compute
  the data term in \eqref{eq:gradient_descent} using shape-gradients
  \citep{herbulot_segmentation_2006}, and ultimately avoid level sets.
We refer the reader to \ref{app:shape_priors} in order to fill in the gap
  between \eqref{eq:gradient_descent} and the following expression:

@@

  \begin{equation}
  sWI = \frac{1}{\sum_i a_i} \sum\limits_i^{N_p} a_i\,\|
  \vec{v}_i - \hat{\vec{v}}_i \|
  \label{eq:swindex}
  \end{equation}
%
  where $\vec{v}_i$ are the locations of the $N_p$ vertices in each $p \in \{1, \dots P\}$
  prior, and $\hat{\vec{v}}_i$ is the location
  recovered corresponding to the vertex $\vec{v}_i$.

@@

\paragraph*{Simulated phantoms}%
\label{sec:digital_phantoms}
We proved the concept on simplistic digital phantoms which we will name after their
  appearance as ``box'', ``ball'', ``L-shape'', and ``gyrus'' (\autoref{fig:phantom},
  boxes A and B).
Using \emph{phantomas} \citep{caruyer_phantomas_2014}, we simulated
  \gls*{t1} (TE/TR=10/1500ms) and \gls*{t2} images (TE/TR=90/5000ms)
  corresponding to each phantom type, with two resolutions each
  ($1.0mm$ and $2.0mm$ isotropic).
Simulations were corrupted with rician noise for a \gls*{snr} of 300.0.
The reference surfaces were extracted at the higher resolution (1.0mm isotropic),
  using \texttt{mri\_tessellate}\footnote{A marching-cubes algorithm shipped with 
  \emph{FreeSurfer}.}.

\begin{figure*}
    \centering
    \resizebox{\textwidth}{!}{%
      \begin{tikzpicture}
        \node[inner sep=10pt, draw=black!40](fig03a) at (0,14.8)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-a}};
        \node[inner sep=10pt, draw=black!40](fig03b) at (0,7.9)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-b}};
        \node[inner sep=10pt, draw=black!40](fig03c) at (0,0)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-c}};
        \node[circle, text=black!75] at (-9.4,17.3) {\Huge \textbf{A}};
        \node[circle, text=black!75] at (-9.4,10.6) {\Huge \textbf{B}};
        \node[circle, text=black!75] at (-9.4,3.5) {\Huge \textbf{C}};
      \end{tikzpicture}
    }%
	\caption{A. The ``cortex'' phantom is a spherical shape with two sulci and an
	  outer crust resembling the cortical folding (left).
	The model is used to generate \gls*{t1} and \gls*{t2} images after warping the
	  contours using a random and plausible transformation $U_{true}^{-1}$ (right).
	B. Visual assessment of the results on the low resolution sets:
	  ``gyrus'' (top-left), ``L-shape'' (top-right), ``ball'' (bottom-left),
	  and ``box'' at (bottom-right).
	In yellow color, the recovered contours after registration are represented.
	Our method showed high accuracy, as demonstrates almost exact location of the registered
    contours with respect to their ground truth position depicted in green.
	Partial volume effect turns segmentation of the sulci a challenging problem with voxel-wise
	  clustering methods, but it is successfully segmented with our method.
	C. Quantitative evaluation of registration error in terms of average Hausdorff distance of
	  surfaces at high (left) and low (right) resolutions, demonstrating that the error is
	  consistently below the voxel size.
	  }\label{fig:phantom}
\end{figure*}

### 2015-03-18 Publications/2015-NeuroImage/05-results.tex
final fixes
improving to second round

\subsection{Proof of concept on digital phantoms}
\label{sec:results_phantom}
Using a simplified subset of the experimental instrumentation described in
  \autoref{sec:experiments_evaluation}, we collected quantitative results on the digital
  phantoms.
A total of 1200 registration experiments (4 phantom types $\times$ 150 random warpings
  $\times$ 2 resolutions) were run.
Parameter settings of every processing step in the experiments are made available along with
  the source code.
For each registration, the averaged Hausdorff distance (see \autoref{sec:experiments_evaluation})
  of the inner and the outer surfaces was computed.
The results showed a consistent and high accuracy, below the image resolution.
\autoref{fig:phantom} (block C) presents the box plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the misregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel size with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals
  (\autoref{tab:ci_phantom}).

\begin{table}
		\centering
		\footnotesize
    \begin{tabular}{llcc}
    Type      & Low-res.  & Hi-res. \\
    \hline
    ``gyrus'' & 0.58-0.60 & 0.18-0.36 \\
    ``L''     & 0.67-0.73 & 0.35-0.39 \\
    ``ball''  & 0.66-0.73 & 0.32-0.41 \\
    ``box''   & 0.68-0.71 & 0.35-0.39 \\
    \hline
    \end{tabular}
    \caption{95\% Confidence Intervals (CIs) were computed using bootstrapping of $10^4$ samples,
      for each phantom type and resolution.
    Taking into account the non-normal distribution of errors, we used the median as location
  		statistic.
    All the CIs were below the corresponding half-voxel size in all the phantom types.}\label{tab:ci_phantom}
\end{table}

@@

\paragraph*{Assessing the segmentation model}\label{sec:res_model_and_metric} %
%
Preliminarily, we investigated the aptness of our segmentation model to play as cost function
  driving the registration process.
First, we evaluated several alternative models to segment the target bivariate images
  (\gls*{fa} and \gls*{adc}).
Based on the observed joint distribution of each (\suppl{section S2}),
  we selected the 6-regions scheme described in \autoref{sec:human_connectome}.

@@

\paragraph*{Evaluation \& cross-comparison}\label{sec:res_cc_evaluation}
%
Finally, we tested the performance of our registration method, and compared it to the
  a standard \gls*{t2b} correction.
We computed the \gls*{swindex} \eqref{eq:swindex} of every surface after correction,
  for both \emph{regseg} and the \gls*{t2b} methods.
Finally, to statistically compare the results, we performed a one-way ANOVA test
  on the warping indices for three specific surfaces of interest
  ($\Gamma_{VdGM}$, $\Gamma_{WM}$, and $\Gamma_{pial}$).
These interfaces are critical in analyses on whole-brain tractography.
The one-way ANOVA evidenced that error distributions were significantly different,
  $p \approx 2 \times 10^{-9}$.
We also reported the 95\% CIs of the \gls*{swindex} for those surfaces.
The aggregated CI for \emph{regseg} was $[1.08 - 1.50] mm$, whereas the competing method
  yielded an aggregated CI of $[2.06 - 2.43] mm$.
The quantitative results of the ANOVA tests and the CIs are presented in \autoref{tab:results_real}.
Visual assessment of one case is shown in \autoref{fig:results_real}, along with the corresponding
  box plots of error distributions by correction method and surface.
Extended visual assessment of the 16 cases is included in the \suppl{section S4}.

>>>>

\subsection{Proof of concept on digital phantoms}
\label{sec:results_phantom}
Using a simplified subset of the experimental instrumentation described in
  \autoref{sec:experiments_evaluation}, we collected quantitative results on the digital
  phantoms.
A total of 1200 registration experiments (4 phantom types $\times$ 150 random warpings
  $\times$ 2 resolutions) were run.
Parameter settings of every processing step in the experiments are made available along with
  the source code.
For each registration, the averaged Hausdorff distance (see \autoref{sec:experiments_evaluation})
  of the inner and the outer surfaces was computed.
The results showed a consistent and high accuracy, below the image resolution.
\autoref{fig:phantom} (block C) presents the violin plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the misregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel size with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals
  (\autoref{tab:ci_phantom}).

\begin{table}
		\centering
		\footnotesize
    \begin{tabular}{lccccc}
    Res. & ``gyrus'' & ``ball''  & ``box''   & ``L''     & Aggreg.    \\\hline
    Low  & 0.59-0.60 & 0.65-0.76 & 0.68-0.71 & 0.67-0.77 & 0.64-0.66  \\
    High & 0.18-0.38 & 0.31-0.45 & 0.34-0.42 & 0.34-0.40 & 0.34-0.38  \\
    \hline
    \end{tabular}
    \caption{95\% Confidence Intervals (CIs) were computed using bootstrapping of $10^4$ samples,
      for each phantom type and resolution.
    Taking into account the non-normal distribution of errors, we used the median as location
  		statistic.
    All the CIs were below the corresponding half-voxel size in all the phantom types.}\label{tab:ci_phantom}
\end{table}

@@

\paragraph*{Assessing the segmentation model}\label{sec:res_model_and_metric} %
%
Preliminarily, we investigated the aptness of our segmentation model to play as cost function
  driving the registration process.
First, we evaluated several alternative models to segment the target bivariate images
  (\gls*{fa} and \gls*{adc}).
Based on the observed joint distribution of each (\suppl{section S4}),
  we selected the 6-regions scheme described in \autoref{sec:human_connectome}.

@@

\paragraph*{Evaluation \& cross-comparison}\label{sec:res_cc_evaluation}
%
Finally, we compared the performance of \emph{regseg} and the standard \gls*{t2b}
  correction.
We computed the \gls*{swindex} \eqref{eq:swindex} of every surface after correction,
  for both \emph{regseg} and the \gls*{t2b} methods.
Finally, to statistically compare the results, we performed a one-way ANOVA test
  on the warping indices for three specific surfaces of interest
  ($\Gamma_{VdGM}$, $\Gamma_{WM}$, and $\Gamma_{pial}$).
These interfaces are critical in analyses of whole-brain tractography.
The one-way ANOVA evidenced that error distributions were significantly different,
  $p \approx 2 \times 10^{-9}$.
We also reported the 95\% CIs of the \gls*{swindex} for those surfaces.
The aggregated CI for \emph{regseg} was $[1.08 - 1.50] mm$, whereas the competing method
  yielded an aggregated CI of $[2.06 - 2.43] mm$.
The quantitative results of the ANOVA tests and the CIs are presented in \autoref{tab:results_real}.
Visual assessment of one case is shown in \autoref{fig:results_real}, along with the corresponding
  violin plots of error distributions by correction method and surface.
Extended visual assessment of the 16 cases is included in the \suppl{section S5}.

### 2015-03-18 Publications/2015-NeuroImage/06-discussion.tex
improving to second round

The hypothesis underpinning our research is that image registration can be reliably performed
  by searching for homogeneous regions in the target image that correspond to precise contours
  from an atlas, or extracted from other image (i.e. a different time step).
We designed \emph{regseg} as a method to correct \gls*{dmri} data for susceptibility
  distortions.
Applications of whole-brain tractography generally include the acquisition of \gls*{t1} images
  to provide with prior information about anatomy with high accuracy.
Our method solves in one single step a joint problem usually addressed in two steps:
  linear registration of \gls*{t1} and \gls*{dmri} to fuse the anatomical information on
  the diffusion data, and distortions correction using extra acquisitions.
Moreover, our method does not require any additional acquisition over the minimal protocol
  that only includes \gls*{t1} and \gls*{dmri} images, a very common situation for
  historical datasets.
Beyond the proposed correction, potential applications of \emph{regseg} are atlas-based
  segmentation and tracking objects in time-series.

\paragraph*{Accuracy tests}
We proved that active contours without edges can be used to successfully drive a
  deformation supported by B-Spline functions in digital phantoms.
We randomly deformed 4 different phantom models mimicking three homogeneous regions,
  showing contrasts generated by a publicly available \gls*{mri} simulator.
After registration with \emph{regseg} we measured the Hausdorff distance between the
  projected contours using the ground-truth warping and our estimation.
We concluded that errors are significantly lower than voxel resolution.
We also assessed the 95\% intervals, yielding an aggregated interval of distances of
  $[0.58 - 0.73]$ in the low resolution phantoms ($2.0mm$ isotropic voxel) and
  $[0.18 - 0.41]$ in the high resolution phantoms ($1.0mm$ isotropic).
These aggregated intervals by resolution take the lowest of the lower boundary
  and the highest of the upper boundary of all CIs regardless the phantom type.
Therefore, we also conclude that the error is bounded and below the half of the
  voxel spacing to illustrate the accuracy of \emph{regseg}.

\paragraph*{Application on real data}
After a preliminary experiment to demonstrate the aptness of our segmentation model
  in section \autoref{sec:res_model_and_metric}, we evaluated \emph{regseg} in
  a real environment, using the experimental framework presented in \autoref{fig:evworkflows}.
We processed 16 subjects from the \gls*{hcp} database with both \emph{regseg}
  and an in-house replication of the \gls*{t2b} correction method.
Our method presented significantly different results (ANOVA tests are reported in
  \autoref{tab:results_real}), and the 95\% CIs revealed the high accuracy of
  \emph{regseg}.
Even though we carefully configured the competing method by using the same algorithm and
  extending the same settings from a widely used software, cross-comparison
  experiments are prone to the so-called instrumentation bias \citep{tustison_instrumentation_2013}.
Therefore, these results do not show that the \emph{regseg} software \emph{is} better than the
  competing method, but they are communicated to proof that \emph{regseg} is a solid option to
  correct \gls*{dmri} data for distortions, and any other potential application.

\paragraph*{Prospects}
Future extensions or improvements of \emph{regseg} are similar to the evolution of existing
  alternative methodologies.
Incorporating knowledge about the distortion by initializing our method with the theory-based
  correction using fieldmaps would be the most profitable.
The main pitfall using fieldmap correction is the challenging problem of unwrapping the
  range of the phase map, as it is clipped in the range $(-\pi, \pi]$.
The most extended solution to this is \emph{prelude} \citep{jenkinson_fast_2003}, and some
  alternatives have been proposed \citep{zhou_reliable_2009,daga_susceptibility_2014} to
  circumvent the inherent unreliability of the method.
The \emph{regseg} algorithm would naturally correct misregistered areas due to errors in
  phase unwrapping, and underestimations introduced by necessary smoothing
  applied to clean up field maps.
This integration has been recently applied in the \emph{b0}-to-\gls*{t2} approach
  by \cite{irfanoglu_susceptibility_2011}, also investigating the usefulness of
  nonregular B-Spline grids.
This is also a potential extension to our deformation model.

>>>>

\paragraph*{Accuracy tests}
The hypothesis underpinning our research was that image registration can be reliably performed
  by searching for homogeneous regions in the target image that correspond to precise contours
  from an atlas, or extracted from other image (i.e. a different time step).
We proved that active contours without edges can be used to successfully drive a
  deformation supported by B-Spline functions using digital phantoms.
We randomly deformed 4 different phantom models mimicking three homogeneous regions,
  showing contrasts generated by a publicly available \gls*{mri} simulator.
After registration with \emph{regseg} we measured the Hausdorff distance between the
  projected contours using the ground-truth warping and our estimation.
We concluded that errors are significantly lower than voxel resolution.
We also assessed the 95\% \gls*{ci}, yielding an aggregated interval of
  $[0.64 - 0.66]mm$ in the low resolution phantoms ($2.0mm$ isotropic voxel) and
  $[0.34 - 0.38]mm$ in the high resolution phantoms ($1.0mm$ isotropic).
Therefore, we also conclude that the error is bounded and below the half of the
  voxel spacing to illustrate the accuracy of \emph{regseg}.

\paragraph*{Application on real data}
We designed \emph{regseg} as a method to correct \gls*{dmri} data for susceptibility
  distortions.
Applications of whole-brain tractography generally include the acquisition of \gls*{t1} images
  to provide with prior information about anatomy with high accuracy.
Our method solves in one single step a joint problem usually addressed in two steps:
  linear registration of \gls*{t1} and \gls*{dmri} to fuse the anatomical information on
  the diffusion data, and correction of distortion using extra acquisitions.
Moreover, our method does not require any additional acquisition over the minimal protocol
  that only includes \gls*{t1} and \gls*{dmri} images, a very common situation for
  historical datasets.
Beyond the proposed correction, potential applications of \emph{regseg} are atlas-based
  segmentation and tracking objects in time-series.

After a preliminary experiment to demonstrate the aptness of our segmentation model
  in section \autoref{sec:res_model_and_metric}, we evaluated \emph{regseg} in
  a real environment, using the experimental framework presented in \autoref{fig:evworkflows}.
We processed 16 subjects from the \gls*{hcp} database with both \emph{regseg}
  and an in-house replication of the \gls*{t2b} correction method.
\emph{Regseg} showed very high accuracy, with an aggregated 95\% \gls*{ci} of $[1.08-1.50]mm$,
  almost below the pixel size of $1.25mm$.
The misregistration error remaining after \emph{regseg} was significantly lower ($p < 0.01$) than the
  error corresponding to the \gls*{t2b} correction, using the one-way ANOVA test
  (\autoref{tab:results_real}).
Visual inspection of all the results (\suppl{section S5}), and the violin plots included in
  \autoref{fig:results_real}, confirmed this intuition.
Even though we carefully configured the \gls*{t2b} method by using the same algorithm and
  same settings from a widely used software, cross-comparison experiments are prone to
  the so-called \emph{instrumentation bias} \citep{tustison_instrumentation_2013}.
Therefore, these results do not show that the \emph{regseg} software \emph{is} better than the
  \gls*{t2b} method, but they are communicated to proof that \emph{regseg} is a solid option to
  correct \gls*{dmri} data for distortions, and any other potential application.

\paragraph*{Prospects}
Future improvements of \emph{regseg} possibly will follow a similar line to the evolution of 
  existing methodologies.
Incorporating knowledge about the distortion by initializing our method with the theory-based
  correction using fieldmaps may be the most cost-effective.
The main pitfall using fieldmap correction is the challenging problem of unwrapping the
  range of the phase map, as it is clipped in the range $(-\pi, \pi]$.
The most extended solution to this is \emph{prelude} \citep{jenkinson_fast_2003}, and some
  alternatives have been proposed \citep{zhou_reliable_2009,daga_susceptibility_2014} to
  circumvent the inherent unreliability of the method.
The \emph{regseg} algorithm would naturally correct misregistered areas due to errors in
  phase unwrapping, and underestimations introduced by necessary smoothing
  applied to clean up field maps.
This integration has been recently applied in the \gls*{t2b} approach
  by \cite{irfanoglu_susceptibility_2011}, also investigating the usefulness of
  nonregular B-Spline grids.
This is also a potential extension to our deformation model.
